#Coursera Reproducible Research 1st Assignment - Activity Monitoring


-------------------------------------------------------------------

Author: Marija Orlova

08.2016

-------------------------------------------------------------------- 


Loading packages that are used during analysis

```{r, echo=TRUE, results = 'hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(data.table)
library(Hmisc)
                  
```

--------------------------------------------------------------------------------

### Loading, reading in the dataset and processing the data

Downloading data on 5-min walk test if there isn't one in working directory

```{r, echo=TRUE}
if (!file.exists("activity.csv")) {
  url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(url, destfile = "activity.zip")
  unzip("activity.zip")
  file.remove("activity.zip")
}
```

Loading "activity.csv" data in R

```{r, echo=TRUE}
a<-read.csv("activity.csv", header=TRUE, sep=",")
```

Changing date variable class to date.

```{r, echo=TRUE}
a$date<-as.Date(a$date, "%Y-%m-%d")
```

-------------------------------------------------------------------------------------------------

## Analysis of total count of steps taken during day (ignoring NAs)

--------------------------------------------------------------------------------------------------

Calculating total steps taken each day.

```{r, echo=TRUE}
stepsbydate<-aggregate(x=a["steps"], by=list(date=a$date), FUN="sum", na.rm=TRUE)
```

### Plotting histogram of the total number of steps taken each day

Plotting histogram of total steps count by day

```{r, echo=TRUE}
h<-ggplot(data=stepsbydate,aes(stepsbydate$steps))+geom_histogram(binwidth = 500)+ labs(title="Total count of staps taken during day", x ="Steps (NAs ingnored)", y= "Frequency")
h
```

### Mean and median number of steps taken each day

Calculating mean and median of steps by each day

```{r, echo=TRUE}
StepMed<-median(stepsbydate$steps, na.rm=TRUE)
StepMean<-mean(stepsbydate$steps, na.rm=TRUE)
```

---------------------------------------------------------------------------------------------------

The mean is `r StepMean` and the median is `r StepMed`.

---------------------------------------------------------------------------------------------------

## Analysis of average count of steps for 5-min time periods

-----------------------------------------------------------------------------------------------------

Calculating average steps count over time intervals
```{r, echo=TRUE}
stepsbyinterval<-aggregate(x=a["steps"], by=list(interval=a$interval), FUN="mean", na.rm=TRUE)
```

Plotting average stapes taken by interval


```{r, echo=TRUE}
g<- ggplot(data=stepsbyinterval, aes(x=interval, y=steps)) + geom_line() + labs(title="Time series plot of the average number of steps taken", x ="5-min interval", y= "Average steps count")
g
```


The 5-minute interval that, on average, contains the maximum number of steps

```{r, echo = TRUE}
maxint<-stepsbyinterval[which.max(stepsbyinterval$steps),"interval"]
```

------------------------------------------------------------------------------------------------------

On average `r maxint` interval contains the maximum number of steps.

------------------------------------------------------------------------------------------------------

## Imputation

------------------------------------------------------------------------------------------------------

### Calculate and report the total number of missing values in the dataset

Claculating how many rows in original dataset contains NA values.

```{r, echo=TRUE}
Narow<-sum(!complete.cases(a))
```

----------------------------------------------

Total number of rows with NAs is `r Narow`.

----------------------------------------------

### Imputing missing data

Creating new data frame
```{r, echo =TRUE}
new<-a
```

Replacing all NAs in steps variable with average count of steps.

```{r, echo =TRUE}
new$steps<-impute(new$steps, mean)
```

Checking if there are any missing values left in data frame.

```{r, echo=TRUE}
na<-sum(!complete.cases(new))
```

---------------------------------------------------

There are `r na` cases left with missing values.

----------------------------------------------------

### Histogram of the total number of steps taken each day 

Calculating totak count of steps taken each day using new imputed data frame

```{r, echo=TRUE}
stepsbydateimp<-aggregate(x=new["steps"], by=list(date=new$date), FUN="sum")
```

Plotting histogram of total steps count each day.


```{r, echo=TRUE}
h2<-ggplot(data=stepsbydateimp,aes(stepsbydateimp$steps))+geom_histogram(binwidth = 500)+ labs(title="Total count of steps taken during day", x ="Steps (Imputed values)", y= "Frequency")
h2
```


###Mean and median total number of steps 

Calculating mean and median of steps by each day using imputed data.

```{r, echo=TRUE}
StepMedimp<-median(stepsbydateimp$steps)
StepMeanimp<-mean(stepsbydateimp$steps)
```

-----------------------------------------------------------------------------------------------

Average steps count ignoring NAs was `r StepMean`, after imputation average is `r StepMeanimp`.

Median of steps count ignoring NAs was `r StepMed`, after imputation median is `r StepMedimp`.

Difference between mean and median after imputation `r StepMedimp-StepMeanimp`.

------------------------------------------------------------------------------------------------

## Average staps during time intervals during weekdays and weekends

------------------------------------------------------------------------------------------------

Creating a factor variable depending on given date is a weekday or weekend day

```{r, echo=TRUE}
new$day<-as.POSIXlt(new$date)$wday
new$wd<-ifelse(new$day %in% c(0,6), "weekend", "weekday")
new$wd<-as.factor(new$wd)
```

Calculating average count of steps taken per 5-min interval depending on date (weekend and weekday).

```{r, echo=TRUE}
bywd<-aggregate(x=new["steps"], by=list(interval=new$interval,wday=new$wd),mean)
```

Creting panel plot.


```{r, echo=TRUE}
z<-ggplot(bywd, aes(x=interval, y=steps)) + geom_line() +facet_grid(.~wday) +labs(title="Average number of steps taken per 5-minute interval across weekdays and weekends", x ="5-min interval", y= "Average steps (imputed values)")
z
```

